---
title: "Added percentages of man-made deaths in EMRO countries, 1990- 2016"
author: "Manal"
date: 2020-01-30
---

In this post, I aim to visualize the man-made causes of death (i.e. conflicts, homicide, suicide & terrorism) in 22 EMRO countries. Here I am using bar plot to do that, but I will do it in a map in another post.

The data is from 1990- 2016 for countries and some regions, and including many causes of deaths.

### Load the libraries

```{r setup, warning=FALSE}
library(tidyverse)
library(knitr)
library(httr)
library(readxl)
library(skimr)
library(wesanderson)
library(RColorBrewer)
library(scales)
```

### Download the data 

This method I followed from one rstudio post, so that this blog post is linked to the data. However, it is also possible to download the data using this [source](https://github.com/rfordatascience/tidytuesday/tree/master/data/2018/2018-04-16) and import it into RStudio directly.

```{r warning = FALSE}
github_link <- "https://github.com/rfordatascience/tidytuesday/raw/master/data/2018/2018-04-16/global_mortality.xlsx"
temp_file <- tempfile(fileext = ".xlsx")
req <- GET(github_link, 
          # authenticate using GITHUB_PAT
           authenticate(Sys.getenv("GITHUB_PAT"), ""),
          # write result to disk
           write_disk(path = temp_file))
mortality <-read_excel(temp_file)
```


### Exploring the data


```{r warning = FALSE}
glimpse(mortality)
```


### Create a subset of WHO EMRO countries (22 countries)
Based on [WHO classification](http://www.emro.who.int/countries.html), I create a subset data to include these 22 countries.

```{r warning = FALSE}
meast <- mortality %>% filter(country %in% c('Afghanistan', 'Djibouti', 'Bahrain', 
                                            'Libya', 'Morocco', 'Egypt', 
                                            'Iran', 'Iraq', 'Pakistan', 'Somalia', 'Jordan', 
                                            'Kuwait', 'Lebanon', 'Oman', 
                                            'Palestine', 'Saudi Arabia', 
                                            'Qatar', 'Syria', 
                                            'United Arab Emirates', 'Yemen', 
                                            'Sudan', 'Tunisia'))

```

### tidy the data
Here I created an interval of 10 years to facilitate visualizing it.

```{r warning = FALSE}
cause_gathered <- meast %>% 
    gather(key = "cause", value = "percent", -year, -country, -country_code)

conflict <- cause_gathered %>% 
                  filter(cause == "Conflict (%)"| cause == "Terrorism (%)"|
                        cause == "Suicide (%)" | cause == "Homicide (%)") %>%
                  separate(cause, c("cause", "(%)"), sep = "( )") %>% 
                  select(country, year, cause, percent) %>%
                  mutate(percent_decimal = percent / 100) %>% 
                  arrange(desc(percent_decimal)) %>%
                  filter(!is.na(percent_decimal)) %>% 
                  mutate(year_class = case_when(between(year, 1990, 2000) ~ "1990-2000",
                                between(year, 2001, 2010) ~ "2001-2010",
                                between(year, 2011, 2016) ~ "2011-2016"))
                   
```

### Graph
Finally, ggplot the data
  
```{r, warning = FALSE}
library(tidytext)  

conflict %>% 
  mutate(country = reorder_within(country, percent_decimal, year_class)) %>% 
  ggplot(aes(x = country,y = percent_decimal, fill = cause)) + 
                        geom_col() +
                        facet_wrap(~ year_class, scales = "free", ncol = 3) +
                        coord_flip() +
                        scale_x_reordered() +
                        theme_minimal() +
                        scale_fill_manual(values = wes_palette("GrandBudapest1", n = 4)) +
                        scale_y_continuous(labels = percent_format()) +
                        labs(y = " ",
                            x = " ",
                            fill = "Cause", 
                            title = "Added percentages of Man-made deaths in EMRO countries, 1990- 2016")



conflict %>% 
  mutate(country = reorder_within(country, percent_decimal, year_class)) %>% 
  ggplot(aes(x = country,y = percent_decimal, fill = cause)) + 
  geom_col() +
  facet_wrap(~ year_class, scales = "free", ncol = 3) +
  coord_flip() +
  scale_x_reordered() +
  theme_minimal() +
  scale_fill_brewer(palette = "Set1") +
  scale_y_continuous(labels = percent_format()) +
  labs(y = " ",
       x = " ",
       fill = "Cause", 
       title = "Added percentages of Man-made deaths in EMRO countries, 1990- 2016")
```

